The error when inferring phylogenies with incipient species by a Birth-Death model

### 2018-04-11 article.tex
Feedback, thanks @rsetienne

  % Two or three sentences to provide a
  % broader perspective, readily comprehensible
  % to a scientist in any discipline, may be included
  % in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
  % by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
  % (The above example is 190 words without the final section, and 250 words with it).
  We conclude that ...

@@

This exploratory research's goal is to produce a 
useful and balanced data set, that quantifies the inference error 
we make from choosing the incorrect prior. In this context, 
we measure the error made on PBD phylogenies, when using a BD prior.

The parameter space the data set explores is inspired on 
[Etienne et al, 2014] for two reasons: (1) the academic
reason of being able to compare results, and (2) the practical
reason that the simulated phylogies will have enough taxa to
be informative, but not too big to be computationally overdemanding.
This research uses a complete superset of those parameter settings.
For the speciation initiation rate, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit, 
extending the set of
values used from one ($0.5$) to three. 
The speciation completion
rates used are $0.1$, $0.3$, $1.0$ and $10^6$ speciation completion
events per (incipient species) lineage per time unit, in which only
the last value is added as a computational proxy for infinity, which
makes the experiment fall back to a Birth-Death model (as speciation
is again instantaneous), allowing to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit,
in which $0.4$ is a new extinction rate.
We investigate these parameters in a factorial fashion, except for
(1) those combinations in which the extinction rate equals or exceeds
the speciation initiation rate, and (2) the difference between
speciation initiation rate and extinction rate exceeds 0.8. This is
done to prevent overly taxon-poor and taxon-rich phylogenies respectively.

@@

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we can use this exact site model as a prior, which we 
know is the truth.
The DNA sequence length and mutation rate was chosen as such to provide one
nucleotide change per 1000 years per lineage. As our time unit is in 
millions of years, and our crown age is 15 million years,
this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K. As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior later.

@@

To visualize the data set, we will produce two figures.
The first figure will display the distribution 
of median nLTT statistics per combination of speciation initiation, speciation
completion and extinction rate. We will use the median nLTT statistic over 
summing nLTT statistic distributions, as we think this will give the cleanest
visualization [NOTE: I am unconvinced, wouldn't a violin plot would
be better?]. We will use the median nLTT statistic, instead of the mean, as
we expect the distribution of nLTT statistics not to be symmetrical. The distribution
of median nLTT statistics is displayed per biological parameter set in a box-plot,
to shown of that distribution as needed, without showing an overly amount of detail
regarding that distribution. 
The second figure will show the distribution of median nLTT statistics per expected
mean duration of speciation. The expected mean duration of parameter can be calculated
with equation 19 of [Etienne \& Rosidell 2012] and is a combination of the three
biological parameters combined to their relevant biological effect. Through these
points, we'll fit a straight line (we have no reason to expect an effect of
higher order) and a locally weighted scatterplot smoothing (LOESS) 
curve, which is a non-parametric fit to expose potential higher order effects.
Additionally, we'll draw a horizontal line to the average nLTT 
value (over all replicates) of those simulations with instantaneous speciation, 
which serves as a baseline of
the inference error made.

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.8\textwidth]{nltt_stats_per_setup.png}
  \caption{
    median nLTT statistic distribution per parameter setting
  }
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.8\textwidth]{nllt_stats_per_mean_dur_spec.png}
  \caption{
    median nLTT statistic distribution per mean duration of speciation
  }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & 1 to 40K \\
    $R_b$                 & RNG seed BEAST2 & 1 to 40K \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters.
  }
  \label{table:simulation_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

  % Two or three sentences to provide a
  % broader perspective, readily comprehensible
  % to a scientist in any discipline, may be included
  % in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced
  % by their inclusion. Under these circumstances, the length of the paragraph can be up to 300 words.
  % (The above example is 190 words without the final section, and 250 words with it).
  We conclude that in general [something], and the effect of sampling is [something].

@@

This exploratory research's goal is to produce a 
useful and balanced data set, that quantifies the inference error 
we make from choosing the incorrect prior. In this context, 
we measure the error made on PBD phylogenies, when using a BD prior.
In brief, we simulate protracted phylogenies using the protracted birth-death process,
from those phylogenies sample a species tree. From that species tree, 
we simulate a DNA sequence alignment. Then we let BEAST infer a phylogeny 
from these alignments using the standard birth-death prior. 
We compare the posterior phylogenies to the species tree using the nLTT statistic.

The parameter space the data set explores is inspired on 
[Etienne et al, 2014] for two reasons: (1) the academic
reason of being able to compare results, and (2) the practical
reason that the simulated phylogies will have enough taxa to
be informative, but not too big to be computationally overdemanding.
This research uses a complete superset of those parameter settings.
For the speciation initiation rate, we'll use $0.1$, $0.5$ and $1.0$ 
speciation initiation events per (good species) lineage per time unit, 
extending the set of
values used from one ($0.5$) to three. 
The speciation completion
rates used are $0.1$, $0.3$, $1.0$ and $10^6$ speciation completion
events per (incipient species) lineage per time unit, in which only
the last value is added as a computational proxy for infinity, which
makes the experiment fall back to a Birth-Death model (as speciation
is again instantaneous), allowing to measure the baseline error.
The extinction rates used are $0.0$, $0.1$, $0.2$ and $0.4$ 
extinction events per (good or incipient) lineage per time unit,
in which $0.4$ is a new extinction rate.
We investigate these parameters in a factorial fashion, except for
(1) those combinations in which the extinction rate equals or exceeds
the speciation initiation rate, and (2) the difference between
speciation initiation rate and extinction rate exceeds $0.8$. This is
done to prevent overly taxon-poor and taxon-rich phylogenies respectively.

@@

From a species tree, we simulate a DNA alignment that has the same history
as the phylogeny. The nucleotides of the DNA alignment follow a JC69
nucleotide substitution model, in which all nucleotide-to-nucleotide transitions
are equally likely. Although this may seem as a simplification, in our Bayesian
inference (see below) we can use this exact site model as a prior, which we 
know is the truth.
The DNA sequence length and mutation rate was chosen as such to provide a
resolution of a 1000 years, that is, to have one expected nucleotide change 
per 1000 years per lineage on average. As our time unit is in 
millions of years, and our crown age is 15 million years,
this equates to a rate of $\frac{1000}{15}$ nucleotide substitutions 
changes per time unit per lineage. To provide for each substitution being unique,
we simulate as much DNA nucleotides as expected mutations, which equals the
nucleotide substitution rate times the crown age, which equals 15K. As this
mutation rate is constant over all branches, it in effect follows a strict 
clock model, which we will specify as the known clock model prior later.

@@

Our results will show the general effect of the biological parameters (speciation initiation, 
speciation completion and extinction rate) using the balanced data set,
and the effect of sampling using the data set conditioned on sampling having an effect.
In both cases, the nLTT statistics distribution will be plotted per
biological parameter set using a violin plot. A violin plot is chosen to
maintain information about the distribution. When displaying the effect of a 
sampling regime, per regime, the nLTT statistics distribution is shown per
biological parameter set.

@@

\begin{figure}[!htbp]
  \includegraphics[width=0.8\textwidth]{fig_nltt_stats_per_setup.png}
  \caption{
    nLTT statistic distribution per biological parameter set, using the
    balanced data set
  }
\end{figure}

\begin{figure}[!htbp]
  \includegraphics[width=0.8\textwidth]{fig_sampling.png}
  \caption{
    nLTT statistic distribution per biological parameter set per sampling
    regime, using the data set conditioned on sampling regime having an effect 
  }
\end{figure}

@@

RJCB and RSE conceived the idea for this experiment and package. 
RJCB created and tested the experiment and package, 
and wrote the first draft of the manuscript. 
RSE contributed substantially to revisions.

% Bibliography
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MEE style
\bibliographystyle{mee}
\bibliography{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}
  \centering 
  \begin{tabular}{l l l}
    \hline
    Parameter             & Description & Values \\
    \hline
    \hline
    $b = b_g = b_i$       & Speciation initiation rate & 0.1, 0.5, 1.0 \\
    $\lambda$             & Speciation completion rate & 0.1, 0.3, 1.0, $10^6$ \\
    $\mu = \mu_g = \mu_i$ & Extinction rate & 0.0, 0.1, 0.2, 0.4 \\
    \hline
    $t_c$                 & Crown age & 15 \\
    $\sigma_c$            & Standard deviation around crown age & 0.001 \\
    $M$                   & Sampling method & Youngest for odd $R_i$, oldest for even $R_i$ \\
    $r$                   & Mutation rate & $\frac{1000}{15}$ \\
    $l_a$                 & DNA alignment length & $15K$ \\
    $f_i$                 & MCMC sampling interval & 1K or more \\
    $R_i$                 & RNG seed incipient tree & 1 to 40K \\
    $R_a$                 & RNG seed alignment simulation & 1 to 40K \\
    $R_b$                 & RNG seed BEAST2 & 1 to 40K \\
    \hline
  \end{tabular}
  \caption{
    Overview of the 12 simulation parameters. Above the horizontal line is 
    the biological parameter set.
  }
  \label{table:simulation_parameters}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
